---
title: "PlotTreeWithODTL"
author: "Bastien Boussau"
date: "31 August 2016"
output: html_document
---

# Plotting gene originations, transfers, duplications and losses.

In this notebook, we plot a species tree decorated with numbers of gene origination, duplication, transfer and loss.

First, I define plotting functions, then I read and prepare the data for plotting, and finally we do the plot.



```{r}
setwd("~/data/plotODTLTree/")
library(ape)
library(RColorBrewer)
library(R.utils)
```

## First, some useful plotting functions and tree handling functions
```{r, cache=F}
library(phytools)
library(plotrix)

# scaling branch lengths
scaler = 1.0

# function to add color bar
# written by Liam J. Revell 2013
# Modified by me to take coordinates and legend
my.add.color.bar.legend<-function(textLegend, leg,cols,title=NULL,lims=c(0,1),digits=1,prompt=FALSE, x, y){
  cex<-0.8
  #Branch colors
  X<-x+cbind(0:(length(cols)-1)/length(cols),1:length(cols)/length(cols))*(leg)
	Y<-cbind(rep(y,length(cols)),rep(y,length(cols))) 		
	lines(c(X[1,1],X[nrow(X),2]),c(Y[1,1],Y[nrow(Y),2]),lwd=4+2,lend=2) 
	for(i in 1:length(cols)) lines(X[i,],Y[i,],col=cols[i],lwd=4,lend=2)
	text(x=x,y=y,round(lims[1],digits),pos=3, cex=cex)
	text(x=x+8*leg/5,y=y,round(lims[2],digits),pos=3, cex=cex)
	if(is.null(title)) title<-"<->"
	text(x=3.5*(2*x+leg)/5,y=y,title,pos=3)
  text(x=X[nrow(X),2],y=Y[nrow(Y),2] ,textLegend,pos=4, cex=cex)

  # Now other annotations: D-T:
  rect(xleft=(2*x+leg/3)/2,ybottom=y-1, xright=(2*x+2*leg/3)/2, ytop=y-0.5, col="bisque4", border=NA)
  text(x=X[nrow(X),2],y=Y[nrow(Y),2]-1,"Duplications",pos=4, cex=cex)
  rect(xleft=(2*x+leg/3)/2,ybottom=y-2, xright=(2*x+2*leg/3)/2, ytop=y-1.5, col="coral3", border=NA)
  text(x=X[nrow(X),2],y=Y[nrow(Y),2]-2,"Transfers",pos=4, cex=cex)

  # Now other annotations: content
  draw.circle(x=(2*x+leg/2)/2,y=y-3, 0.1, col="darkolivegreen1")
  text(x=X[nrow(X),2],y=Y[nrow(Y),2]-3,"Content",pos=4, cex=cex)
  # Now other annotations: origination
  draw.circle(x=(2*x+leg/2)/2,y=y-4, 0.1, col="darkorchid3")
  text(x=X[nrow(X),2],y=Y[nrow(Y),2]-4,"Originations",pos=4, cex=cex)
}

##################################
##################################

# functions to add edge labels
# found in ape
# Modified by me to tolerate width argument, and remove the frame 
my.BOTHlabels <- function (text, sel, XX, YY, adj, frame, pch, thermo, pie, piecol, 
    col, bg, horiz, width, height, ...) 
{
    if (missing(text)) 
        text <- NULL
    if (length(adj) == 1) 
        adj <- c(adj, 0.5)
    if (is.null(text) && is.null(pch) && is.null(thermo) && is.null(pie)) 
        text <- as.character(sel)
    frame <- match.arg(frame, c("rect", "circle", "none"))
    args <- list(...)
    CEX <- if ("cex" %in% names(args)) 
        args$cex
    else par("cex")
    if (frame != "none" && !is.null(text)) {
        if (frame == "rect") {
            width <- strwidth(text, uniT = "inches", cex = CEX)
            height <- strheight(text, uniT = "inches", cex = CEX)
            if ("srt" %in% names(args)) {
                args$srt <- args$srt%%360
                if (args$srt == 90 || args$srt == 270) {
                  tmp <- width
                  width <- height
                  height <- tmp
                }
                else if (args$srt != 0) 
                  warning("only right angle rotation of frame is supported;\n         try  `frame = \"n\"' instead.\n")
            }
            width <- xinch(width)
            height <- yinch(height)
            xl <- XX - width * adj[1] - xinch(0.03)
            xr <- xl + width + xinch(0.03)
            yb <- YY - height * adj[2] - yinch(0.02)
            yt <- yb + height + yinch(0.05)
            rect(xl, yb, xr, yt, col = bg)
        }
        if (frame == "circle") {
            radii <- 0.8 * apply(cbind(strheight(text, uniT = "inches", 
                cex = CEX), strwidth(text, uniT = "inches", 
                cex = CEX)), 1, max)
            symbols(XX, YY, circles = radii, inches = max(radii), 
                add = TRUE, bg = bg)
        }
    }
    if (!is.null(thermo)) {
        parusr <- par("usr")
        if (is.null(width)) {
            width <- CEX * (parusr[2] - parusr[1])
            width <- if (horiz) 
                width/15
            else width/40
        }
        if (is.null(height)) {
            height <- CEX * (parusr[4] - parusr[3])
            height <- if (horiz) 
                height/40
            else height/15
        }
        if (is.vector(thermo)) 
            thermo <- cbind(thermo, 1 - thermo)
        thermo <- if (horiz) 
            width * thermo
        else height * thermo
        if (is.null(piecol)) 
            piecol <- rainbow(ncol(thermo))
        if(height > 0) {
          xl <- XX - width/2 + adj[1] - 0.5
          xr <- xl + width
          # yb <- YY - height/2 + adj[2] - 0.5 # My mod
          yb <- YY + adj[2] - 0.5
          yt <- yb + height
        }
       else {
          xl <- XX - width/2 + adj[1] - 0.5
          xr <- xl + width
          # yb <- YY - height/2 + adj[2] - 0.5 # My mod
          yb <- YY - adj[2] - 0.5
          yt <- yb + height
       }
        if (horiz) {
            rect(xl, yb, xl + thermo[, 1], yt, border = NA, col = piecol[1])
            for (i in 2:ncol(thermo)) rect(xl + rowSums(thermo[, 
                1:(i - 1), drop = FALSE]), yb, xl + rowSums(thermo[, 
                1:i]), yt, border = NA, col = piecol[i])
        }
        else {
            rect(xl, yb, xr, yb + thermo[, 1], border = NA, col = piecol[1])
            for (i in 2:ncol(thermo)) rect(xl, yb + rowSums(thermo[, 
                1:(i - 1), drop = FALSE]), xr, yb + rowSums(thermo[, 
                1:i]), border = NA, col = piecol[i])
        }
        s <- apply(thermo, 1, function(xx) any(is.na(xx)))
        xl[s] <- xr[s] <- NA
        rect(xl, yb, xr, yt, border = NA)
    }
    if (!is.null(pie)) {
        if (is.vector(pie)) 
            pie <- cbind(pie, 1 - pie)
        xrad <- CEX * diff(par("usr")[1:2])/50
        xrad <- rep(xrad, length(sel))
        XX <- XX + adj[1] - 0.5
        YY <- YY + adj[2] - 0.5
        for (i in seq_along(sel)) {
            if (any(is.na(pie[i, ]))) 
                next
            floating.pie.asp(XX[i], YY[i], pie[i, ], radius = xrad[i], 
                col = piecol)
        }
    }
    if (!is.null(text)) 
        text(XX, YY, text, adj = adj, col = col, ...)
    if (!is.null(pch)) 
        poinT(XX + adj[1] - 0.5, YY + adj[2] - 0.5, pch = pch, 
            col = col, bg = bg, ...)
}

my.edgelabels<-function (text, edge, adj = c(0.5, 0.5), frame = "rect", pch = NULL, 
    thermo = NULL, pie = NULL, piecol = NULL, col = "black", 
    bg = "lightgreen", horiz = FALSE, width = NULL, height = NULL, 
    date = NULL, ...) 
{
    lastPP <- get("last_plot.phylo", envir = .PlotPhyloEnv)
    if (missing(edge)) {
        sel <- 1:dim(lastPP$edge)[1]
        subedge <- lastPP$edge
    }
    else {
        sel <- edge
        subedge <- lastPP$edge[sel, , drop = FALSE]
    }
    if (lastPP$type == "phylogram") {
        if (lastPP$direction %in% c("rightwarD", "leftwarD")) {
            XX <- (lastPP$xx[subedge[, 1]] + lastPP$xx[subedge[, 
                2]])/2
            YY <- lastPP$yy[subedge[, 2]]
        }
        else {
            XX <- lastPP$xx[subedge[, 2]]
            YY <- (lastPP$yy[subedge[, 1]] + lastPP$yy[subedge[, 
                2]])/2
        }
    }
    else {
        XX <- (lastPP$xx[subedge[, 1]] + lastPP$xx[subedge[, 
            2]])/2
        YY <- (lastPP$yy[subedge[, 1]] + lastPP$yy[subedge[, 
            2]])/2
    }
    if (!is.null(date)) 
        XX[] <- max(lastPP$xx) - date
    print("XX: ")
    print(XX)
    print("YY: ")
    print(YY)
    piecolD = "bisque4" 
    piecolT = "coral3"  
    print(">>>XX")
    print(XX)
    print(">>>YY")
    print(YY)
    print(">>>width")
    print(width)
    print(">>>height")
    print(height)
    maxHeight <- max(abs(height))
    #We want the max height to be 0.6.
    heightNorm <-  0.6 * height / maxHeight  
    sapply(1:length(XX),function(i){
      if (height[i] > 0) { 
        print("height > 0")
        rect(XX[i]-abs(heightNorm[i]/2), YY[i]+0.1, XX[i]+abs(heightNorm[i]/2), YY[i]+heightNorm[i]+0.1, border = NA, col=piecolD )
      }else{ 
        print("height < 0")
        rect(XX[i]-abs(heightNorm[i]/2), YY[i]-0.1, XX[i]+abs(heightNorm[i]/2), YY[i]+heightNorm[i]-0.1, border = NA, col=piecolT ) 
      }
    })
}


my.edgelabelsSeparateDAndT<-function (Dvalues, Tvalues, text, edge, adj = c(0.5, 0.5), frame = "rect", pch = NULL, 
    thermo = NULL, pie = NULL, piecol = NULL, col = "black", 
    bg = "lightgreen", horiz = FALSE, width = NULL, height = NULL, 
    date = NULL, ...) 
{
    lastPP <- get("last_plot.phylo", envir = .PlotPhyloEnv)
    if (missing(edge)) {
        sel <- 1:dim(lastPP$edge)[1]
        subedge <- lastPP$edge
    }
    else {
        sel <- edge
        subedge <- lastPP$edge[sel, , drop = FALSE]
    }
    if (lastPP$type == "phylogram") {
        if (lastPP$direction %in% c("rightwards", "leftwards")) {
            XX <- (lastPP$xx[subedge[, 1]] + lastPP$xx[subedge[, 
                2]])/2
            YY <- lastPP$yy[subedge[, 2]]
        }
        else {
            XX <- lastPP$xx[subedge[, 2]]
            YY <- (lastPP$yy[subedge[, 1]] + lastPP$yy[subedge[, 
                2]])/2
        }
    }
    else {
        XX <- (lastPP$xx[subedge[, 1]] + lastPP$xx[subedge[, 
            2]])/2
        YY <- (lastPP$yy[subedge[, 1]] + lastPP$yy[subedge[, 
            2]])/2
    }
    if (!is.null(date)) 
        XX[] <- max(lastPP$xx) - date
    print("XX: ")
    print(XX)
    print("YY: ")
    print(YY)
    piecolD = "bisque4" #brewer.pal(1, "YlOrRd")
    piecolT = "coral3"  #brewer.pal(1, "YlOrRd")
    print(">>>XX")
    print(XX)
    print(">>>YY")
    print(YY)
    maxHeight <- max(abs(c(Dvalues, Tvalues)))
    #We want the max height to be 0.6.
    DvaluesNorm <-  0.6 * Dvalues / maxHeight 
    TvaluesNorm <- 0.6 * Tvalues / maxHeight 
    sapply(1:length(XX),function(i){
      rect(XX[i]-abs(DvaluesNorm[i]/2), YY[i]+0.1, XX[i]+abs(DvaluesNorm[i]/2), YY[i]+DvaluesNorm[i]+0.1, border = NA, col=piecolD )
      rect(XX[i]-abs(TvaluesNorm[i]/2), YY[i]-0.1, XX[i]+abs(TvaluesNorm[i]/2), YY[i]-TvaluesNorm[i]-0.1, border = NA, col=piecolT ) 
    })
}





########################################################
########################################################
plotDTLThermoOri<-function(treeDTL, xDTL,  mode=c("edges","tips","nodes"),palette="rainbow",legend=TRUE,xlims=NULL,...){
  ############################
  #DTL matrix
  # For all nodes
  allLosDTL <- xDTL$L
  allDupsDTL <- xDTL$D
  allTrasDTL <- xDTL$Tra
  allOriDTL <- xDTL$Os
    
  sizer <- 0.4
  
  #For internal nodes
  dupDTL = xDTL$D[which(xDTL$internal==TRUE)]
  traDTL = xDTL$Tra[which(xDTL$internal==TRUE)] 
  losDTL = xDTL$L[which(xDTL$internal==TRUE)]
  conDTL = sqrt(xDTL$copies[which(xDTL$internal==TRUE)])
  oriDTL = sqrt(xDTL$Os[which(xDTL$internal==TRUE)])

  
  #For tips
  duptDTL = xDTL$D[which(xDTL$internal==FALSE)]
  tratDTL = xDTL$Tra[which(xDTL$internal==FALSE)] 
  lostDTL = xDTL$L[which(xDTL$internal==FALSE)]
  contDTL = sqrt(xDTL$copies[which(xDTL$internal==FALSE)])
  oritDTL = sqrt(xDTL$Os[which(xDTL$internal==FALSE)])

    print("oritDTL")
    print(oritDTL)
  
############################
# Useful variables
  maxdup <- max(xDTL$D)
  maxtra <- max(xDTL$Tra)  
  maxlos <- max(xDTL$L)
  maxcon <- sqrt(max(xDTL$copies))
  allLos <- allLosDTL

############################
# Normalization

  d.valueDTL = 0.3+allDupsDTL/maxdup
  t.valueDTL = 0.3+allTrasDTL/maxtra
  d.valueDTLd_t = (allDupsDTL - allTrasDTL) / max(allDupsDTL - allTrasDTL)
  
  con.valueDTL=sizer*(0.0+conDTL/maxcon)
  cont.valueDTL=sizer*(0.0+contDTL/maxcon)

  # We size the originations same as the counts
  ori.valueDTL=sizer*(0.0+oriDTL/maxcon)
  orit.valueDTL=sizer*(0.0+oritDTL/maxcon)
  

  pies <- rep(1, length(con.valueDTL)+1)
  piest <- rep(1, length(cont.valueDTL)+1)
  
  ############################
  # Starting the plotting stuff
  
  # Defining colors
	if(hasArg(tol)){tol<-list(...)$tol
	}else{ tol<-1e-6}
	if(palette=="heat.colors") cols<-heat.colors(n=1000)
	if(palette=="gray") cols<-gray(1000:1/1000)
	if(palette=="rainbow")	cols<-rainbow(1000,start=0.7,end=0) # blue->red
	if(is.null(xlims)) xlims<-range(allLos)+c(-tol,tol)
	breaks<-0:1000/1000*(xlims[2]-xlims[1])+xlims[1]
	whichColor<-function(p,cols,breaks){
		i<-1
		while(p>=breaks[i]&&p>breaks[i+1]) i<-i+1
		cols[i]
	}

  colorsDTL<-sapply(allLosDTL, whichColor, cols=cols, breaks=breaks)
  
  piecolD = "bisque4" #brewer.pal(1, "YlOrRd")
  piecolT = "coral3" #brewer.pal(1, "YlOrRd")
  piecolCon = "darkolivegreen1" #brewer.pal(1, "YlOrRd")
  piecolOri = "darkorchid3"
############################
par (mar=c(0,0,0,0),mfrow=c(1,1))

  # DTL Tree plotting
	par(lend=2)
  print ("XXX TREE 1")
	plot.phylo(treeDTL,no.margin=TRUE,edge.width=3,edge.color=colorsDTL,label.offset=0.02*max(nodeHeights(treeDTL)),lend=2,new=FALSE, y.lim=c(1,length(treeDTL$tip.label)*1.2), cex=0.6)
	
my.edgelabelsSeparateDAndT(d.valueDTL, t.valueDTL, thermo=pies, horiz=F, col = piecolD, piecol=piecolD, pos=4, adj = c(0.45, 0.85), cex = d.valueDTLd_t, bg=rgb(0,0,0,0), frame="none", height= d.valueDTLd_t/10, width=d.valueDTLd_t/10) # Dup-T values as a thermo

#Content pie plots
  # Internal nodes
  nodelabels(pie=pies, horiz=T, col = piecolCon, piecol=piecolCon, adj = c(0.5, 0.5), cex = con.valueDTL, height= 0, width=0.5) # Content values  as a pie
  # Tip nodes
  tiplabels(pie=piest, horiz=T, col = piecolCon, piecol=piecolCon, adj = c(0.5, 0.5), cex = cont.valueDTL, height= 0, width=0.5) # Content values  as a pie

#Ori pie plots
  # Internal nodes
  nodelabels(pie=pies, horiz=T, col = piecolOri, piecol=piecolOri, adj = c(0.5, 0.5), cex = ori.valueDTL, height= 0, width=0.5) # Content values  as a pie
  # Tip nodes
  tiplabels(pie=piest, horiz=T, col = piecolOri, piecol=piecolOri, adj = c(0.5, 0.5), cex = orit.valueDTL, height= 0, width=0.5) # Content values  as a pie
  
  
    	if(legend==TRUE&&is.logical(legend)) legend<-round(0.3*max(nodeHeights(treeDTL)),2)
	if(legend){
		if(hasArg(title)) title<-list(...)$title
		else title<-NULL
		if(hasArg(digits)) digits<-list(...)$digits
		else digits<-1
		my.add.color.bar.legend("Losses", legend,cols,title,xlims,digits,prompt=FALSE, 0.0393, length(treeDTL$tip.label)*1.15)
	}
}

#################################
plotAllVariables <- function(dDTL, labelsDTL, treeDTL, name){ 
  # Reorganize the rows so that the tree and the DTL tables are in the same order
  tmp <- dDTL[unlist(sapply(labelsDTL$label,function(z)which(dDTL$branch==z))),]
  rownames(tmp) <- labelsDTL$label
  tmp_dtl <- tmp[sapply(treeDTL$edge[,2],function(z)which(rownames(tmp)==labelsDTL$label[z])),]

   plotDTLThermoOri(treeDTL, tmp_dtl, name, palette="rainbow")

}
```


## Now, a function to reformat and normalize a table that has just been read.

```{r, echo=T}
reformatTable <- function(dt, labelsdt){ 
  #Renaming some columns
dt$branch <- dt$name
dt$Tra <- dt$Ts
dt$D <- dt$Ds
dt$L <- dt$Ls
dt$branch <- as.character(dt$branch)
dt$normD <- dt$D/(dt$D+dt$Tra)
dt$normT <- dt$Tra/(dt$D+dt$Tra)
dt$normL <- dt$L/(dt$D+dt$Tra)
#Change in copy number minus duplications
dt$deltaNorm <- dt$copies - dt$D +dt$L
# Fraction of new genes due to transfer
dt$transNormbyGenSize <- dt$Tra / dt$copies

#Now, we reorder d according to node ID
dt<-merge(dt, labelsdt, by.x="branch", by.y="label")
dt$ID<-as.numeric(as.character(dt$ID))
dt<-dt[order(dt$ID),]
#Internal node or not?
dt$internal<-lapply(dt$branch, function (x) {if (length(grep("[a-zA-Z]",as.character(x))) == 1) {return (FALSE)} else {return (TRUE)}})
return (dt)
}

```

## Reading the tree

```{r, echo=T}
# First, we read the tree
treeDTL<-read.tree("testData/test.tree")
treeDTL$edge.length = scaler*treeDTL$edge.length


# Link between internal node iD in R and the node names/iD used in the table
tipNames<-treeDTL$tip.label
nodeNames<-treeDTL$node.label
tipAndNodeNames <- c(tipNames, nodeNames)

labels <- rbind(cbind(1:length(treeDTL$tip.label),treeDTL$tip.label),cbind((length(treeDTL$tip.label)+1):(length(treeDTL$tip.label) + length(treeDTL$node.label)),treeDTL$node.label))
colnames(labels) <- c("ID","label")
labelsDTL <- as.data.frame(labels)
```


## Reading the data table

```{r, echo=T}
dDTL<-read.table("testData/test.txt", h=T, comment.char="")

dDTL<-reformatTable(dDTL, labelsDTL)

```

## Getting the link between the short names and the full names

```{r}
labnam <- read.table("testData/species_namesOK.txt", fill=F, h=T, sep=",", stringsAsFactors=F, quote="")
labnam$Code <- as.character(labnam$Code)
labnam$LongName <- as.character(labnam$LongName)
splittedNames <- lapply(labnam$LongName, function(x) strsplit(x, " ")[[1]])
labnam$LongName<- lapply(splittedNames, function(x) paste(x[1], " ", (x[2]), " ", sep=""))
rownames(labnam) <- labnam$Code
treeDTL$tip.label<-labnam[treeDTL$tip.label,2]
```

## Altering the gene content estimates so that they are corrected for the genes that have been lost in all descendant genomes

```{r}
correct <- read.table("testData/test.ext", h=F)
correct$V1 <- as.character(correct$V1)
names(correct) <- c("branch", "correction")
dDTL<-merge(dDTL, correct, by.x="branch", by.y="branch", all.x=T)
dDTL$correction[which(is.na(dDTL$correction))] <- 1
dDTL$copies <- dDTL$copies / dDTL$correction
```

## Plotting along the tree:

```{r, echo=T, fig.width=6, fig.height=7}

print (summary(dDTL$Os))


pdf(file="plotODTL.pdf", width=7, height=14 )
plotAllVariables(dDTL, labelsDTL, treeDTL, "Genome content and col")
dev.off()

```

